Catching fire: Undefined behaviour for equational reasoning of lazy languages with strictness annotations

### 2021-01-21 paper/thesis.tex
nuke large sections that are outdated

% 1
We propose a new style of semantics
for non-strict languages that provide access to many optimizations that are
currently unsound for lazy languages.
% 2 
Lazy languages, while providing equational reasoning, are unable to reason
effectively about \emph{strictness annotations}. On the other hand, most strict
languages are impure, and thus unable to provide equational reasoning due to
the presence of side effects.
% 3
Research on optimizing non-strict languages has focused on optimizing away
non-strictness with demands, and efficient lowering of strictness onto modern
hardware.
% 4
We take a different approach: we propose to use undefined behaviour judiciously
while defining the semantics of our language to enable many key optimizations that
are possible in the strict world, that are not possible in the non-strict world.
We model \emph{divergent computations as undefined behaviour}, giving the
compiler far more freedom to optimize and reorder a mix of strict and non-strict
code.
% 5 
We first consider the various deficiencies of introducing an explicit "bang pattern"
or "strictness annotation" as is done within the Glasgow Haskell compiler today.
Then we demonstrate how setting divergent computations to be undefined behaviour 
cleanly enables the compiler to optimize these programs.
% 6
This modelling of divergence as undefined behaviour enables us to write an
\emph{optimizing} compiler that can reason about a mix of strictness and non
strictness. Furthermore, it enables us to modularly add non-strictness into
an existing strict language. We demonstrate this with a prototype optimizing
functional language called Lizzy. Lizzy uses MLIR as a backend. We introduce
a new dialect \lz in MLIR for modularly expressing non-strictness.  As a benchmark,
we consider a shootout between GHC's optimization of vector and matrix
computations, and \lz's ability to optimize the same computations.
% (http://www.easterbrook.ca/steve/2010/01/how-to-write-a-scientific-abstract-in-six-easy-steps/)
\end{abstract}

@@

% \grosser{Always state your contributions explicitly: (A) this makes it
% easy for the reader to understand what novelty is presented, and (B)
% these contributions help you to focus. In particular, the objectie of
% the remaining paper should be to support the claims stated here.
% }
Our contributions are:                                          
\begin{itemize}
\item A formal semantics for bang patterns in the Haskell language.
\item A study of where bang patterns inhibit GHC's ability to reason about strictness.
\item A novel yet simple extension to a lazy language semantics that remove this
      inability of GHC (and lazy languages in general) to reason about strictness.
\item An implementation of a prototype lazy language (Lizzy) that chooses this
      approach pervasively.
\item A new dialect (\lz) for MLIR to encode these semantics, and provide modular
      non-strictness.
\end{itemize}

@@

\section{Strictness Annotations: The Bad And The Ugly}

We first motivate why strictness annotations inhibit the compiler
from being able to optimize programs. Consider the program in
\autoref{fig:lazy-loop}, where \texttt{out} declares
a binding \raw{x = loop} and proceeds to not use it. This allows GHC
to optimize out the program into \raw{out = 42}. On the
other hand, consider  \autoref{fig:strict-loop} where \raw{out}
declares  \emph{strict} binding \raw{!x  = loop} and proceeds to not
use it.

\begin{lstlisting}[language=haskell,caption=unoptimized \raw{out} and optimized \raw{outOptimized}]
{-# LANGUAGE BangPatterns #-}
loop, out, out' :: Int
loop = loop;
out = let x = loop in 42
outOptimized = 42
main :: IO ()
main = do print out; print out'
\end{lstlisting}

\begin{lstlisting}[language=haskell, caption=unoptimized \raw{outStrict} with an incorrect optimization in \raw{outStrictOptimized}]
{-# LANGUAGE BangPatterns #-}
loop, out, out' :: Int
loop = loop;
outStrict = let !x = loop in 42
outStrictOptimized = 42 -- INCORRECT TRANSFORM
main :: IO ()
main = do print out; print out'
\end{lstlisting}
% \caption{An example of how strictness annotations break equational reasoning}

\begin{lstlisting}[language=haskell, caption=Original haskell program]
{-# LANGUAGE BangPatterns #-}
out b = let large = unknown 10 in 
  if False then large else 42
out' b = let !large = unknown  10 in 
  if False then large else 42
main :: IO ()
main = do print out; print out'
\end{lstlisting}

\begin{lstlisting}[language=Haskell, caption=Correct optimization that is incorrect with a bang pattern]
{-# LANGUAGE BangPatterns #-}
loop, out, out' :: Int
loop = loop;
out b= if b then 
  let large = unknown 10 in large else 42
-- | incorrect. If `large` had been divergent, 
-- then `out` no longer
-- exhibits divergence
out' b = if b then 
  let !large = unknown 10 in large else 42

main :: IO ()
main = do print out; print out'
\end{lstlisting}
% \caption{An example of how strictness annotations disallow floating a \texttt{let}
%          down into the use-site, inhibiting optimization}

\section{Undefined Behaviour For Divergence?}

As we have shown, the existence of bang patterns obviates equational reasoning.
We are now in a bind: we can choose to forego bang patterns, thereby
regaining equational reasoning with a loss of performance. This is not a 
very reasonable suggestion, as bang patterns are critical for controlling
evaluation ordering, and ensuring that we do not build up thunks.

We first propose a simple, outrageous, solution:  Ban all divergence.
We define divergence to be \emph{undefined behaviour}. In essence, this
allows the compiler to assume that \emph{no value} is divergent, for if a
value were divergent, then \emph{anything} can happen. We analyze the
merits and deficiencies of this proposal, and then provide a more
refined version of the above solution.

% \begin{figure}[ht]
% \begin{minted}{hs}
% {-# LANGUAGE BangPatterns #-}
% loop, out, out' :: Int
% outStrict = let !x = loop in 42
% -- ==If x were not divergent==, 
% --   it is safe to set out' to 42.
% -- ==If x were divergent==
% --   It is undefined behaviour. Anything can
% --   happen, including evaluating out' to 42.
% outStrictOptimized = 42 
% main :: IO ()
% main = do print out; print out'
% \end{minted}
% \caption{An example of how defining divergence as UB allows continued
%          equational reasoning with bang patterns.}
% \label{fig:strict-loop-ub}
% \end{figure}

Consider the program in \autoref{fig:strict-loop}. Here, according
to our semantics, we define \emph{divergence as undefined behaviour}.
Interpreted according to these semantics, here's how we legally
allow the transformation \raw{outStrictOptimized = 42}:

\begin{itemize}
    \item Since \raw{x} is forced, we must evaluate \raw{loop} into weak head normal form.
    \item \raw{loop} does not have a weak head normal form. This is a divergent computation.
    \item Since we have divergence, \emph{anything} can happen [we can refine $\bot$ into a
    value that is greater than $\bot$].
    \item We choose to refine $\bot$ into $42$ as $\bot \sqsubset 42$.
\end{itemize}

%\begin{figure}
%\begin{minted}{hs}
\begin{lstlisting}[language=haskell, caption=An example of how defining divergence as UB is too agressive]
{-# LANGUAGE BangPatterns #-}
-- | never evaluates to a value.
main = putStrLn "42"; main
mainOptimized = return () 
 -- | legal if divergence is UB.
\end{lstlisting}
%\end{minted}
%\label{fig:io-loop}
%\caption{An example of how defining divergence as UB is too agressive}
%\end{figure}

On the other hand, this definition is \emph{far too agressive}. Consider
the program in \autoref{fig:io-loop}. Here, main prints $42$ forever,
never returning the final \raw{IO ()} that it is expected to. Thus,
we can \emph{optimize this program} into \raw{mainOptimized = return ()}
if divergence is undefined behaviour. This is not what we want! We wish
to allow programs which \emph{make progess without halting}, while
still being able to exploit undefined behaviour. Thus, we must change our
semantics to adapt to this need.

\section{Progressive operational semantics}

Naively speaking, one maybe tempted to summarize our approach as
"make divergence undefined". While this is appealing, this is both too
strong, and too naive. For example, such a definition would allow us to
optimize away programs that \emph{progress} without \emph{terminating}.
A tiny example of such a program, which prints ``42'' forever is
given in \autoref{fig:print-forever}

% \begin{figure}[ht]
% \begin{minted}{hs}
\begin{lstlisting}[language=haskell, caption=A program that makes progress without terminating.  We wish to preserve such program semantics]
action :: IO (); action = putStrLn "42";
main :: IO (); main = action >> main
\end{lstlisting}
% \end{minted}
% \label{fig:print-forever}
% \caption{A program that makes progress without terminating.  We wish to preserve such program semantics}
% \end{figure}

% \begin{figure}[ht]

%\begin{subfigure}[b]{0.48\textwidth}
%\begin{minted}{hs}
\begin{lstlisting}[language=haskell]
{-# LANGUAGE BangPatterns #-}
loop :: Int; loop = loop
ex1 = let x = loop in 42
ex2 = let !x = loop in 42
ex3 = let !x = loop in x
-- | New program
{-# LANGUAGE BangPatterns #-}
loop :: Int; loop = loop
ex1 = 42
ex2 = 42
ex3 = let !x = loop in x
ex3' = loop -- equivalent semantics
\end{lstlisting}
% \end{minted}
% \subcaption{Maxmially transformed programs with our proposed semantics}
% \end{subfigure}
% \caption{Programs that exhibit a range of behviours, clarifying what values we do and do not eliminate}
% \end{figure}

We call our semantics \emph{progressive}, which tries to make as much progress
as possible before getting stuck at a stuck term.
\section{Formal Semantics Of Bang Patterns in Haskell}

There is currently no formal definition of bang patterns in the Haskell language.
The semantics are defined based on how the Glasgow Haskell Compiler chooses
to manipulate these strictness annotations. So wherever we have previously argued
that it would be ``incorrect'' for GHC to perform these transforms, we have been
arguing from the position of:

\begin{itemize}
\item This is what GHC does today, and that is the current state of definition for bang patterns.
\item Were GHC to make the above transformation, we believe it would violate the
      intuitions that many people have about how GHC \emph{ought} to transform code.
      \sid{I asked bgamari who works on GHC full-time, he said that GHC does not perform the above transform most likely.
           Similarly, other friends (carter, davean) who commit into GHC also said likely not.}
\end{itemize}

To begin our discussion, we first formalize the semantics of bang patterns.
\textbf{TODO: fill in semantics here.}
\section{\lz: Its Syntax, Type system, semantics}
\subsection{A high level overview of \mlir}

@@

% =========================
\vspace{20em}

@@

\vspace{10em}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
\texttt{\%x = constructor(@Constructor, \%v1, ..., \%vm)} \\
\texttt{\%y = case \%x} \\
\texttt{      [\@C1 -> \{\string^entry1(\%z11, ..., \%z1n1): ...  \}]} \\
\texttt{      [...]} \\
\texttt{      [\@Ci -> \{\string^entry1(\%zi1, ..., \%zini): ...  \}]} \\
}{
\texttt{Inline \string^entryi with \%zik = \%vk}
}
} % end footnotesize
\subcaption{Case of known constructor: remove indirection}
\end{minipage}

\vspace{10em}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Inline a function call}
\end{minipage}
\caption{Local rewrites performed to eliminate laziness (1)}
\end{figure*}

\begin{figure*}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Outline recursive call of constructor that is immediately unwrapped}
\end{minipage}
% OutlineRecursiveApEagerOfConstructorPattern

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Outline pattern matching branches on a function input}
% OutlineCaseOfFnInput
\end{minipage}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Outline return of constructor}
% outline return of constructor
\end{minipage}

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
FOO
}{
BAR
}
} % end footnotesize
\subcaption{Peel constructor from case}
% peel constructor from case
\end{minipage}

%% % \begin{minted}{text}
%% % Inlining
%% % \end{minted}
\caption{Local rewrites performed to eliminate laziness (2)}
\end{figure*}

>>>>

% 1
We propose a new style of semantics
for non-strict languages that provide access to many optimizations that are
currently unsound for lazy languages.
% 2 
Lazy languages, while providing equational reasoning, are unable to reason
effectively about \emph{strictness annotations}. On the other hand, most strict
languages are impure, and thus unable to provide equational reasoning due to
the presence of side effects.
% 3
Research on optimizing non-strict languages has focused on optimizing away
non-strictness with demands, and efficient lowering of strictness onto modern
hardware.
% 4
We take a different approach: we propose to use undefined behaviour judiciously
while defining the semantics of our language to enable many key optimizations that
are possible in the strict world, that are not possible in the non-strict world.
We model \emph{divergent computations as undefined behaviour}, giving the
compiler far more freedom to optimize and reorder a mix of strict and non-strict
code.
\end{abstract}

@@

% \grosser{Always state your contributions explicitly: (A) this makes it
% easy for the reader to understand what novelty is presented, and (B)
% these contributions help you to focus. In particular, the objectie of
% the remaining paper should be to support the claims stated here.
% }
Our contributions are:                                          
\begin{itemize}
\item A new dialect (\lz) for MLIR to encode these semantics, and provide modular
      non-strictness.
\item A translator from STG to \lz that exhibits the ability to 
      eliminate laziness and leverage MLIR's loop optimisation  framework.
\item A rewrite driven strategy for worker/wrapper that interleaves 
     worker/wrapper analysis along with worker/wrapper rewriting, by phrasing
     the worker/wrapper problem as an outlining/inlining problem.
\end{itemize}

@@

\textbf{TODO: fill in semantics here.}
\section{\lz: Its Syntax, Type system, semantics}
\subsection{A high level overview of \mlir}

@@

% =========================
\vspace{15em}

@@

\vspace{15em}

\begin{tabular}{ll}
\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
\texttt{\%x = constructor(@Constructor, \%v1, ..., \%vm)} \\
\texttt{\%y = case \%x} \\
\texttt{      [\@C1 -> \{\string^entry1(\%z11, ..., \%z1n1): ...  \}]} \\
\texttt{      [...]} \\
\texttt{      [\@Ci -> \{\string^entry1(\%zi1, ..., \%zini): ...  \}]} \\
}{
\texttt{Inline \string^entryi with \%zik = \%vk}
}
} % end footnotesize
\subcaption{Case of known constructor: remove indirection}
\end{minipage}
&

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
\texttt{data C = MkC(V) }\\
\texttt{@f(\%inc: C)} \\
\texttt{ case inc} \\
\texttt{   [C inv -> } \\
\texttt{     \%inv = extract(@MKC, \%inc) : V}\\
\texttt{     \%w = ... : V; \%wc = construct(@MKC, w) : C} \\
\texttt{     \%rec = apEager(@f, wc)]}\\
}{
\texttt{@frec(\%inv: V)}\\
\texttt{ \%inv = extract(@MKC, \%inc) : V}\\
\texttt{ \%w = ... : V; \%wc = construct(@MKC, w) : C} \\
\texttt{ \%rec = apEager(@f, wc)}\\
\texttt{@f(\%inc: C)} \\
\texttt{ \%inv = extract(\%inc) : V}\\
\texttt{ apEager(@frec, inv)}\\
}
} % end footnotesize
\subcaption{Outline recursive call of constructor that is immediately unwrapped}
\end{minipage}
% OutlineRecursiveApEagerOfConstructorPattern
\end{tabular}

\vspace{15em}
\begin{tabular}{ll}
\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
}{
BAR
}
} % end footnotesize
\subcaption{Outline pattern matching branches on a function input}
% OutlineCaseOfFnInput
\end{minipage}
&

\begin{minipage}[t][1cm][b]{0.5\textwidth}
{\footnotesize
\inference{
\texttt{data C = MKC(V) }\\
\texttt{@f(...)}\\
\texttt{  ...}\\
\texttt{  \%out = constructor(C, \%w)}\\
\texttt{  lz.return (\%out)}\\
}{
\texttt{data C = MKC(V) }\\
\texttt{@finner(...)} \\
\texttt{ ...} \\
\texttt{ lz.return \%w} \\
\texttt{ } \\
\texttt{@f(...)}\\
\texttt{ \%w = call \%f(...)}\\
\texttt{ \%out = constructor(C, \%w)}\\
\texttt{ lz.return (\%out)}
}
} % end footnotesize
\subcaption{Outline return of constructor}
% outline return of constructor
\end{minipage}
\end{tabular}
%% 
%% 
%% \begin{minipage}[t][1cm][b]{0.5\textwidth}
%% {\footnotesize
%% \inference{
%% FOO
%% }{
%% BAR
%% }
%% } % end footnotesize
%% \subcaption{Peel constructor from case}
%% % peel constructor from case
%% \end{minipage}
\caption{Local rewrites performed to eliminate laziness (2)}
\end{figure*}
